% ============================================================
%  acknowledgements.tex — Remerciements
% ============================================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

We would like to express our sincere gratitude to our
supervisors, Mme~Malika Mehdi-Silhadi and Mme~Mekahlia,
for their time, guidance, and patience throughout this
project.
Their feedback consistently redirected the work when it
lost focus, and their encouragement made the difficult
stretches of the experimentation manageable.

We thank the Department of Bioinformatics at USTHB for
the academic environment and the resources that supported
this research.

We are grateful to the authors of the breast DCE-MRI
dataset --- Alves and Boulogne at Radboud University
Medical Centre --- for making their data publicly
available, and to the Flower framework team for their
open-source federated learning tools.

Finally, we thank our families.
This thesis was written over many late evenings and
weekends; without their understanding and support,
it would not have been completed.

The authors acknowledge the use of a generative AI
assistant (Claude, Anthropic) during the preparation of
this thesis. AI assistance was used for the generation of
LaTeX markup, and the review of technical writing.
All claims, experimental results, code, and interpretations
are the original work of the authors and were verified
independently. The use of AI tools was limited to
drafting and notation support and does not affect the
originality of the research contributions.
