\chapter*{General Introduction}
\addcontentsline{toc}{chapter}{General Introduction}

% ============================================================
% PROBLEMATIQUE — General Introduction
% ============================================================
\section*{Research Problem and Questions}

The convergence of three concurrent realities motivates
the research conducted in this thesis.
First, breast cancer molecular subtype classification from
DCE-MRI is a clinically consequential task: the subtype
determines whether a patient receives endocrine therapy,
cytotoxic chemotherapy, or targeted anti-HER2 treatment,
and early, accurate classification directly influences
survival outcomes.
Second, the datasets required to train reliable deep
learning classifiers for this task are distributed across
hospital institutions that cannot freely share patient
imaging data: clinical records are locked in isolated
institutional silos whose incompatible systems and
restricted access policies make centralised collection
impractical without a distributed learning approach.
Third, the dataset available at any single institution is
severely class-imbalanced, with the rarest subtype
representing fewer than 7\% of cases, placing a hard
ceiling on what any single-site model can achieve.

Federated learning offers a principled resolution to the
first two realities by enabling multi-institutional model
training without raw data exchange.
However, the third reality introduces a challenge that
federated learning alone does not resolve: when data are
both scarce and unevenly distributed across institutions,
the standard Federated Averaging algorithm converges
toward a globally biased model that performs well on
majority classes and poorly on the clinical minorities
that are most diagnostically challenging.
The central research problem of this thesis is therefore:

\begin{center}
\textit{How can federated learning be configured and
optimised to classify breast MRI molecular subtypes
accurately across privacy-preserving hospital silos,
given severe class imbalance and non-identically
distributed data?}
\end{center}

This problem is decomposed into three specific research
questions that structure the experimental evaluation:

\begin{enumerate}

  \item What classification performance ceiling does a
  centralised deep learning system achieve on breast MRI
  molecular subtype prediction under severe class
  imbalance, and how does a clinically motivated binary
  reformulation address the fundamental data scarcity
  constraint imposed by rare subtypes?

  \item How much classification performance is lost when
  training is distributed across three simulated hospital
  clients with non-IID data partitioning under the FedAvg
  aggregation strategy, and what is the nature of the
  privacy--performance trade-off in this setting?

  \item Can SCAFFOLD's control-variate correction
  mechanism reduce the convergence gap caused by
  statistical heterogeneity in a three-hospital federated
  breast MRI setting, and to what extent does
  convergence-aware aggregation recover the performance
  lost to client drift?

\end{enumerate}

\section*{Contributions}

This thesis makes the following original contributions
to the federated medical imaging literature.

\begin{itemize}

  \item A complete three-stage training pipeline combining
  ConvNeXt-Nano with Gated Attention Multiple Instance
  Learning, Label-Distribution-Aware Margin loss, Adaptive
  Sharpness-Aware Minimization, Stochastic Weight
  Averaging, and Classifier Retraining, designed
  specifically for severely imbalanced breast DCE-MRI
  classification on a single consumer-grade GPU.

  \item A systematic empirical comparison of FedAvg and
  SCAFFOLD aggregation strategies under three degrees of
  non-IID data heterogeneity across simulated hospital
  clients, using a clinically motivated binary
  reformulation of the molecular subtype task.

  \item A proof-of-concept radiomic feature fusion
  experiment combining 256-dimensional deep MIL embeddings
  with 59 hand-crafted PyRadiomics features, demonstrating
  the complementary information available from explicit
  texture descriptors on whole-breast DCE-MRI volumes
  without tumour segmentation masks.

\end{itemize}

The experimental validation is conducted on breast
Dynamic Contrast-Enhanced MRI (DCE-MRI) molecular
subtype classification, a domain selected deliberately
on four grounds.
First, breast DCE-MRI is among the most legally and
ethically sensitive medical imaging modalities:
volumetric MRI studies contain biometric and genetic
implications that make cross-institutional sharing
particularly constrained under both international and
Algerian data protection law.
Second, the publicly available Duke Hospital dataset
\cite{saha2018,alves2023breastmri} presents a class
imbalance ratio of 10.4:1, making it one of the most
challenging multi-class imbalance benchmarks in medical
imaging FL research and a rigorous test for the
proposed optimisation techniques.
Third, the molecular subtype label has direct,
high-stakes treatment consequences: the binary
Luminal versus Non-Luminal decision governs whether
a patient receives endocrine therapy or cytotoxic
chemotherapy, making classification errors clinically
costly in a federated deployment context.
Fourth, it is the only large-scale publicly available
breast DCE-MRI collection with molecular subtype
labels, ensuring that results are reproducible and
comparable.
The optimisation contributions of this thesis,
including the three-stage imbalance-aware training
pipeline and the SCAFFOLD aggregation strategy, are
general and transferable to other healthcare FL
scenarios beyond breast MRI.

% ============================================================
